---
title: "Wastewater_HAV Analysis"
author: ""
date: today
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
  #pdf: default
editor: visual
theme: cosmo
execute:
  echo: true
  warning: false
  message: false
---

# Introduction

This report presents the analysis of surveillance data to assess:

\-

\-

\-

```{r}
#| warning: false
#| message: false
#| echo: false

library(glmmTMB)
library(broom.mixed)
library(dplyr)
library(knitr)
library(tidyverse)
library(lubridate)
library(janitor)
library(lme4)
library(performance)
library(sjPlot)
library(gt)
```

# Results

## Descriptive Statistics

```{r}
#| echo: false
#| message: false
#| warning: false

library(tidyverse)
library(lubridate)
library(janitor)
# Load dataset
waste <- read.csv("data/Wastewatergeneral-HAV.csv")
waste <- clean_names(waste) # data clean
```

```{r}
#| warning: false
#| message: false
#| echo: false
# Clean Outcome Variable
# colSums(is.na(waste)) # Missing

# Clean Outcome Variable
waste <- waste %>%
  mutate(
    hav_result = trimws(hav_result),
    hav_result = tolower(hav_result)
  )
# Check
#unique(waste$hav_result)

# Create Analysis Dataset
waste_analysis <- waste %>%
  filter(hav_result %in% c("positive", "negative"))

#unique(waste_analysis$hav_result)

# Create Binary Outcome
waste_analysis <- waste_analysis %>%
  mutate(
    hav_result_bin = ifelse(hav_result == "positive", 1, 0)
  )

# Verify it
#table(waste_analysis$hav_result)
#table(waste_analysis$hav_result_bin)

# Compute Positivity Rate
# mean(waste_analysis$hav_result_bin) * 100
# Summary Table
su=waste_analysis %>%
  count(hav_result) %>%
  mutate(
    percent = n / sum(n) * 100
  )

# Convert Date Properly for Trend Analysis
waste_analysis$sample_collection_date <- mdy(waste_analysis$sample_collection_date)

```

### Positive Rate and summary table

-   Positive rate of HAV result is `r round((mean(waste_analysis$hav_result_bin) * 100), 3)`.
-   As shown in Table @tbl-sumt, out of `6329` valid waste-water tested samples, `1993` were hepitetis A virus (HAV) positive test result and `4336` HAV_test negative result.

```{r}
#| warning: false
#| message: false
#| echo: false

# Compute Positivity Rate
prhav=mean(waste_analysis$hav_result_bin) * 100

```

```{r}
#| warning: false
#| message: false
#| echo: false
#| label: tbl-sumt
#| tbl-cap: "Summary Statistics"
# Summary Table
sumt=waste_analysis %>%
  count(hav_result) %>%
  mutate(
    percent = n / sum(n) * 100
  )
knitr::kable(sumt)
```

### Monthly positive Trend (Epidemic Curve)

As @fig-moncases indicates, the monthly positivity trend of

```{r}
#| warning: false
#| message: false
#| echo: false

# Monthly Positivty Trend
monthly_trend <- waste_analysis %>%
  mutate(month = floor_date(sample_collection_date, "month")) %>%
  group_by(month) %>%
  summarise(
    positivity = mean(hav_result_bin) * 100,
    n = n()
  )


```

```{r}
#| warning: false
#| message: false
#| echo: false
#| label: fig-ptrend
#| fig-cap: "Monthly Positivity Trend"
ggplot(monthly_trend,
      aes(x = month, y = positivity)) +
  geom_line() +
  geom_point()

```

### Positive by province and district

-   Positivity trend by provincee is @tbl-pbp

-   Positivity trend by district is @tbl-pbd

```{r}
#| warning: false
#| message: false
#| echo: false
#| label: tbl-pbp
#| tbl-cap: "Positivity Trend by Provinces"
# Positivity by Province
pbp=waste_analysis %>%
  group_by(site_province) %>%
  summarise(
    positivity = mean(hav_result_bin) * 100,
    total_tests = n()
  ) %>%
  arrange(desc(positivity))

knitr::kable(pbp)


```

```{r}
#| warning: false
#| message: false
#| echo: false
#| label: tbl-pbd
#| tbl-cap: "Positivity Trend by District"

# Positivity by District
pbd=waste_analysis %>%
  group_by(district_name ) %>%
  summarise(
    positivity = mean(hav_result_bin) * 100,
    total_tests = n()
  ) %>%
  arrange(desc(positivity))

knitr::kable(pbd)
```

### Laboratory Quality Indicators

```{r}
#| warning: false 
#| message: false 
#| echo: false  

# Failed test rate 
failed_rate=mean(waste$hav_result == "failed") * 100  
# Missing outcome rate 
#missing_rate=mean(waste$hav_result == "") * 100
```

-   Failed test rate of the laboratory is `r round(failed_rate, 3)`.
-   Missing HAV test rate of the laboratory `r round(mean(waste$hav_result == "") * 100, 3)`.

## Multilevel Logistic Regression

### Intraclass Correlation for site_province Variation

```{r}
#| warning: false
#| message: false
#| echo: false
### Multilevel Logistic Regression
# A. Site Province Level
model_province <- glmer(
  hav_result_bin ~ (1 | site_province),
  data = waste_analysis,
  family = binomial()
)

# B. District level
model_district <- glmer(
  hav_result_bin ~ (1 | district_name),
  data = waste_analysis,
  family = binomial()
)

# C. Site name level
model_site_name <- glmer(
  hav_result_bin ~ (1 | site_name),
  data = waste_analysis,
  family = binomial()
)

# Random intercept models for site_name

model1 <- glmer(
  hav_result_bin ~ site_province + (1 | site_name),
  data = waste_analysis,
  family = binomial()
  )

model2 <- glmer(
  hav_result_bin ~ site_province + district_name + (1 | site_name),
  data = waste_analysis,
  family = binomial()
)

```

An intraclass correlation coefficient (ICC) of 0.033 means that about 3.3% of the total variance in the HAV is attributable to differences between site_province, while about 96.7% is due to differences within groups (individual-level variation).

```{r}
#| warning: false
#| message: false
#| echo: false
performance::icc(model_province)
```

### Intraclass Correlation for District_name Variation

An intraclass correlation coefficient (ICC) of 0.054 means that about 5.4 % of the total variance in the HAV is attributable to differences between district level.

```{r}
#| warning: false
#| message: false
#| echo: false
performance::icc(model_district)
```

### Intraclass Correlation for site_name Variation

An intraclass correlation coefficient (ICC) of 0.095 means that about 9.5 % of the total variance in the HAV is attributable to differences between district level.

Overall:

```{r}
#| warning: false
#| message: false
#| echo: false
performance::icc(model_site_name)
```

-   A mixed-effects logistic regression with a random intercept for province and district showed that variation across them (province-level ICC ≈ 3.3%, and district level ICC ≈ 5.4%), indicating some clustering of the outcome within provinces and districts.

-   Relatively the variation is higher in site_name level (ICC ≈ 9.5%)

-   @tbl-mlrim indicates, model4; the three level mixed effects model better fits the data.

```{r}
#| warning: false
#| message: false
#| echo: false
#| label: tbl-mlrim
#| tbl-cap: "Multilevel models assessment and comparisons"

AIC(model1, model2)
```

## Hierarchical Analysis based on the selected Modlel

From the hierarchical modeling @tbl-m3lr, a generalized linear mixed model with a random intercept for site was fitted to assess geographic predictors of HAV positivity. Site-name clustering was low (ICC ≈ 0.095). Significant district-level differences were observed, with Buffalo City MM associated with lower odds of HAV positivity and Frances Baard DM associated with substantially higher odds. Province-level effects were not statistically significant after adjustment. The model exhibited rank deficiency due to collinearity between district and province variables.

```{r}
#| warning: false
#| message: false
#| echo: false
#| label: tbl-m3lr
#| tbl-cap: "Three level Multilevel Logistic Regression"
#######################################
### Multilevel Logistic Regression

summary(model2)
# knitr::kable(round(summary(model2)$coefficients, 3))
 #vc2=as.data.frame(VarCorr(model2))
#knitr::kable(round(vc2, 3), caption = "Random effects Model for Site Names")

```

### Odds Ratios

Odds ratio of @tbl-m3lrOR

```{r}
#| warning: false
#| message: false
#| echo: false
#| label: tbl-m3lrOR
#| tbl-cap: "Three level Multilevel Estimated Odds Ratio (OR)"
#######################################
# Odds Ratio
OR <- exp(fixef(model2))

knitr::kable(
  data.frame(
    Term = names(OR),
    Odds_Ratio = round(OR, 3)
  )
)
```

### Visualization of Model Results

Visualization plot @fig-mr3

```{r}
#| warning: false
#| message: false
#| echo: false
#| label: fig-mr3
#| fig-cap: "Visualization of three-level Model Results"
sjPlot::plot_model(
  model2,
  type = "est",
  show.values = TRUE,
  value.offset = 0.3
)
```

### Model Diagnostics

-   Residual diagnostics indicated mild deviations from the expected residual distribution, although no significant overdispersion or strong outlier effects were detected. Remaining structure in residuals suggests that additional predictors, particularly temporal effects, may improve model fit.

```{r}
#| warning: false
#| message: false
#| echo: false
# Model Diagnostics
library(DHARMa)

isSingular(model2)

sim_res <- simulateResiduals(model2)
plot(sim_res)

testDispersion(sim_res)

testOutliers(sim_res)

# Multicolinarity
# library(car)
# vif(lm(hav_result_bin ~ site_province + district_name,
# data = waste_analysis))

```

## Time Trend Analysis

-   In this section, outcome of the dataset aggregated by month only. We can check it by `str(monthly_count`).

```{r}
#| warning: false
#| message: false
#| echo: false

# Monthly Aggregate Counts
monthly_counts <- waste_analysis %>%
  mutate(month = floor_date(sample_collection_date, "month")) %>%
  group_by(month) %>%
  summarise(
    positive_cases = sum(hav_result_bin),
    total_tests = n(),
    positivity = mean(hav_result_bin, na.rm = TRUE) * 100,
    .groups = "drop"
  )

# Create Time Variable
monthly_counts <- monthly_counts  %>%
  arrange(month) %>%
  mutate(
    time = dense_rank(month)
  )

# Drop Missing values
monthly_counts <- monthly_counts %>%
  filter(!is.na(month))
```

Monthly hav case over time @fig-moncases shows a clear shift from very low hav case counts in 2020/2021 to 2022 to a major surge beginning in early 2024, followed by a gradual decline with continued variability through 2025-2026.

-   Structural change beginning around 2023.A major epidemic wave in 2024.
-   Transition in to a lower but still persistent transmission level afterward.
-   2020/21 - 2022: cases stayed near zero, with occasional small spikes.
-   2023: a gradual upward trend begins, with several moderate increases reaching around 30-40 cases. Early 2024: a sharp outbreak occurs, with cases rapidly climbing above 100.
-   Mid 2024: the highest peak appears around 125-30 cases.
-   Late 2024-2025: cases decline but remain elevated compared with earlier years, fluctuating roughly between 35 and 75.

<!-- The post-peak fluctuations suggest the system may not yet have fully stabilized.-->

```{r}
#| warning: false
#| message: false
#| echo: false
#| label: fig-moncases
#| fig-cap: "Monthly hav case over time"
# Plot cases over time
library(ggplot2)
ggplot(monthly_counts, aes(x= month, y = positive_cases)) + 
  geom_line(color = "blue") +
  geom_point() +
  labs(title = "monthly cases over time",
       x = "month",
       y = "Number of cases")
```

### Add Moving Average line

-   The monthly positivity trend demonstrated substantial temporal variation, with a marked increase during 2024 followed by a gradual decline. The moving-average curve suggested a nonlinear temporal pattern.

```{r}
#| warning: false
#| message: false
#| echo: false

library(dplyr)
library(ggplot2)
library(zoo)
# Add a Moving Average Trend Line
monthly_counts <- monthly_counts %>%
  arrange(month) %>%
  mutate(MA_3 = rollmean(positive_cases, 3, fill = NA, align = "center"))

ggplot(monthly_counts, aes(x = month)) +
  geom_line(aes(y = positive_cases), color = "black") +
  geom_point(aes(y = positive_cases), size = 1.5) +
  geom_line(aes(y = MA_3), color = "blue", linewidth = 1.2) +
  labs(
    title = "Monthly Positive Cases by adding MA line",
    y = "positive_cases"
  ) +
  theme_minimal()
```

### Detect Outbreaks

-   Points doted by red line detected as outbreak at the time of 2024.

```{r}
#| warning: false
#| message: false
#| echo: false
# Detect Outbreak
mean_cases <- mean(monthly_counts$positive_cases, na.rm = TRUE)
sd_cases <- sd(monthly_counts$positive_cases, na.rm = TRUE)

threshold <- mean_cases + 2 * sd_cases

monthly_counts$outbreak <- monthly_counts$positive_cases > threshold

# Highlight outbreak Monthly
ggplot(monthly_counts, aes(month, positive_cases)) +
  geom_line() +
  geom_point(aes(color = outbreak), size = 2) +
  scale_color_manual(values = c("black", "red")) +
  theme_minimal()
```

################################################################# 

## Time Series Analysis after areal Aggregation

-   In this section, outcome of the dataset aggregated by month, site_province, district_name, and site_name. We can check it by `str(aggreg_counts`).

-   All analysis and results are similar with previous section.

-   The only difference is the data structure.

```{r}
#| warning: false
#| message: false
#| echo: false
# Dataset B: 
# Aggregate Counts
aggreg_counts <- waste_analysis %>%
  mutate(month = floor_date(sample_collection_date, "month")) %>%
  group_by(site_province, district_name, site_name, month) %>%
  summarise(
    positive_cases = sum(hav_result_bin),
    total_tests = n(),
    positivity = mean(hav_result_bin, na.rm = TRUE) * 100,
    .groups = "drop"
  )
```

-   Creating areal data from the original dataset grouping by (month, site_province, district_name, and site_name)

```{r}
#| warning: false
#| message: false
#| echo: false

# Create Time Variable
aggreg_counts <- aggreg_counts  %>%
  arrange(month) %>%
  mutate(
    time = dense_rank(month)
  )
# Drop Missing values
model_data <- aggreg_counts %>%
  filter(!is.na(site_province), !is.na(district_name), !is.na(site_name), !is.na(month))
```

### Trend Analysis for the areal data plot

Monthly hav case over time @fig-moncases1 shows a clear shift from very low hav case counts in 2020/2021 to 2022 to a major surge beginning in early 2024, followed by a gradual decline with continued variability through 2025-2026.

-   Structural change beginning around 2023.A major epidemic wave in 2024.
-   Transition in to a lower but still persistent transmission level afterward.
-   2020/21 - 2022: cases stayed near zero, with occasional small spikes.
-   2023: a gradual upward trend begins, with several moderate increases reaching around 30-40 cases. Early 2024: a sharp outbreak occurs, with cases rapidly climbing above 100.
-   Mid 2024: the highest peak appears around 125-30 cases.
-   Late 2024-2025: cases decline but remain elevated compared with earlier years, fluctuating roughly between 35 and 75.

```{r}
#| warning: false
#| message: false
#| echo: false
#| label: fig-moncases1
#| fig-cap: "Monthly hav case over time-1"
# Plot cases over time
library(ggplot2)
ggplot(aggreg_counts, aes(x= month, y = positive_cases)) + 
  geom_line(color = "blue") +
  geom_point() +
  labs(title = "monthly cases over time-1",
       x = "month",
       y = "Number of cases")
```

### Add Moving Average line for the areal data plot

```{r}
#| warning: false
#| message: false
#| echo: false

library(dplyr)
library(ggplot2)
library(zoo)
# Add a Moving Average Trend Line
aggreg_counts <- aggreg_counts %>%
  arrange(month) %>%
  mutate(MA_3 = rollmean(positive_cases, 3, fill = NA, align = "center"))

ggplot(aggreg_counts, aes(x = month)) +
  geom_line(aes(y = positive_cases), color = "black") +
  geom_point(aes(y = positive_cases), size = 1.5) +
  geom_line(aes(y = MA_3), color = "blue", linewidth = 1.2) +
  labs(
    title = "Monthly Positive Cases",
    y = "positive_cases"
  ) +
  theme_minimal()
```

### Detect Outbreaks

```{r}
#| warning: false
#| message: false
#| echo: false
# Detect Outbreak
mean_cases <- mean(aggreg_counts$positive_cases, na.rm = TRUE)
sd_cases <- sd(aggreg_counts$positive_cases, na.rm = TRUE)

threshold <- mean_cases + 2 * sd_cases

aggreg_counts$outbreak <- aggreg_counts$positive_cases > threshold

# Highlight outbreak Monthly
ggplot(aggreg_counts, aes(month, positive_cases)) +
  geom_line() +
  geom_point(aes(color = outbreak), size = 2) +
  scale_color_manual(values = c("black", "red")) +
  theme_minimal()
```

```{r}
#| warning: false
#| message: false
#| echo: false

# Check seasonality and Trend
# Identify start time
# head(monthly_counts$month)
# min(monthly_counts$month, na.rm = TRUE)
# Convert to time series objects
#ts_pos_cases <- ts(monthly_counts$positive_cases,
 #              start = c(2020, 7),
  #             frequency = 12)

# plot(ts_pos_cases)

# Check Seasonality
ts_pos_cases1 <- ts(aggreg_counts$positive_cases, frequency = 12)
# decomp <- stl(ts_data, s.window = "periodic")
# plot(decomp)
```

### Decomposition Plot

The decomposition plot revealed a pronounced seasonal pattern with consistent periodic oscillations throughout the study period. The trend component demonstrated a gradual increase followed by substantial fluctuations, indicating nonstationary behavior and possible nonlinear temporal dynamics. Residual variability increased over time, suggesting heteroscedasticity and the presence of unexplained irregular fluctuations or outbreak-like events. Overall, the findings support the inclusion of seasonal and nonlinear temporal terms in subsequent modeling approaches.

-   The time series decomposition plot separates the observed series in to four parts.

    -   Data (observed series)

    -   Seasonal component

    -   Trend component

    -   Residual

-   Observed data: presence both trend and seasonality.

    -   There is fluctuations over time.

    -   Variability appear to increase after around time 50 -60.

    -   peak become become in later period

-   Seasonal component: seasonality appears consistent.

    -   Very regular oscillations

    <!-- -->

    -   Stable repeating cycles across the whole time period

    <!-- -->

    -   Seasonal effect magnitude remains relatively constant

-   Trend component: nonlinear increasing trend in the series

    -   Slow increase initially

    <!-- -->

    -   Strong upward movement around time ≈ 50–70

    <!-- -->

    -   Then fluctuating but remaining elevated

-   Residual component: the unexplained part after removing trend and seasonality

    -   Residuals centered around zero

    <!-- -->

    -   Residual variability increases over time

```{r}
#| warning: false
#| message: false
#| echo: false
# Decomposition
# decom=decompose(ts_pos_cases, s.window = "periodic")
decom1=stl(ts_pos_cases1, s.window = "periodic")
plot(decom1)
```

```{r}
#| warning: false
#| message: false
#| echo: false
# Examine Autocorrelation 
# Autocorrelation Function (ACF)
#acf(ts_pos_cases,
 #   main = "ACF of Monthly Cases")
```

```{r}
#| warning: false
#| message: false
#| echo: false
# Examine Autocorrelation 
# Partial Autocorrelation Function (PACF)
#pacf(ts_pos_cases,
 #    main = "PACF of monthly cases")
```

```{r}
#| warning: false
#| message: false
#| echo: false
# Plot 
ggplot(aggreg_counts, aes(month, positive_cases)) +
  geom_line(color = "black", linewidth = 0.7) +
  geom_smooth(se = FALSE, color = "blue", linewidth = 1.2) +
  labs(
    title = "Monthly Positive Cases Over Time",
    subtitle = "Observed monthly counts with smoothed trend",
    x = "Month",
    y = "Positive Cases"
  ) +
  theme_classic(base_size = 14)
```

### Count (Poisson) Time Series Modeling

Trend of monthly positive cases @fig-ttcases

```{r}
#| warning: false
#| message: false
#| echo: false
#| label: fig-ttcases
#| fig-cap: "Monthly Positive Cases"
# Visualize Time Trend

library(ggplot2)

ggplot(aggreg_counts,
       aes(x = month,
           y = positive_cases)) +
  geom_line() +
  geom_point() 
  #labs(
   # title = "Monthly Positive Cases",
  #  x = "Month",
   # y = "Positive Cases"
  #)
```

#### Poisson Time Series Model

Although the outcome consists of count data suitable for Poisson-based modeling, the observed autocorrelation, trend, and seasonal dependence indicate that simple Poisson regression would inadequately account for temporal dependence. Time-series count models incorporating autoregressive and seasonal structures, such as SARIMA, GLARMA, INGARCH, or Negative Binomial GAM frameworks, are likely more appropriate for robust inference and forecasting.

Baseline poisson model is @tbl-bpm

```{r}
#| warning: false
#| message: false
#| echo: false
#| label: tbl-bpm
#| tbl-cap: "Baseline Poisson Time-Series Model"
# Fit Baseline Poisson Time-Series Model
pois_model <- glm(
  positive_cases ~ time,
  family = poisson(),
  data = model_data
)

# summary(pois_model)

knitr::kable(round(summary(pois_model)$coefficients, 3))

```

##### Assessment

Overdespersion of poisson regression model @tbl-ovdtest

```{r}
#| warning: false
#| message: false
#| echo: false
#| label: tbl-ovdtest
#| tbl-cap: "Overdispersion Test for Poisson Time-Series Model-1"
# Check Overdispersion
library(performance)

od <- check_overdispersion(pois_model)

od_df <- data.frame(
  Dispersion_Ratio = round(od$dispersion_ratio, 3),
  Pearson_Chi2 = round(od$chisq_statistic, 3),
  P_Value = round(od$p_value, 3)
)

knitr::kable(
  od_df)
```

#### Negative binomial time-series model

Negative binomial time-series model @tbl-nbtsm

```{r}
#| warning: false
#| message: false
#| echo: false
#| label: tbl-nbtsm
#| tbl-cap: "Negative Binomial Time-Series Model"
#### Fit Negative Binomial Time-Series Model
library(MASS)

nb_model <- glm.nb(
  positive_cases ~ site_province + district_name + site_name + time,
  data = model_data
)

# summary(nb_model)

knitr::kable(round(summary(nb_model)$coefficients, 3))

```

Negative binomial time series model by adding seasonality @tbl-nbseason

```{r}
#| warning: false
#| message: false
#| echo: false
#| label: tbl-nbseason
#| tbl-cap: "Seasonal Negative Binomial Time-Series Model"
# Add Seasonality
model_data$month_num <-
  month(model_data$month)

nb_seasonal <- glm.nb(
  positive_cases ~ site_province + district_name + site_name + time + factor(month_num),
  data = model_data
)

# summary(nb_seasonal)

knitr::kable(round(summary(nb_seasonal)$coefficients, 3))
```

Negative binomial time series models by adding offset @tbl-nboffset

```{r}
#| warning: false
#| message: false
#| echo: false
#| label: tbl-nboffset
#| tbl-cap: "Negative Binomial Time-Series Model- Adding offset"
# Add Population Offset
nb_offset <- glm.nb(
  positive_cases ~ site_province + district_name + site_name + time +
    offset(log(total_tests)),
  data = model_data
)

# summary(nb_offset)

knitr::kable(round(summary(nb_seasonal)$coefficients, 3))
```

### Mixed effects NB

GLMM negative binomial time series models is give as @tbl-nbglmm

```{r}
#| warning: false
#| message: false
#| echo: false
#| label: tbl-nbglmm
#| tbl-cap: "Mixed Effects Negative Binomial Time-Series Model"
# GLMM (Mixed Effect NB Model)
nb_glmm <- glmer.nb(positive_cases ~ time + (1 | site_province), data = model_data)

summary(nb_glmm)
knitr::kable(round(summary(nb_glmm)$coefficients, 3))
```

##### Negative Binomial Mixed MOdel

```{r}
#| warning: false
#| message: false
#| echo: false
# Negative Binomial Mixed Model
library(glmmTMB)
nb_lmixed = glmmTMB(
  positive_cases ~ + time + factor(month_num) + (1 | site_name),
  family = nbinom2,
  data = model_data
)

summary(nb_lmixed)

```

##### Comparison of negative binomial models @tbl-nbcomparison

```{r}
#| warning: false
#| message: false
#| echo: false
#| label: tbl-nbcomparison
#| tbl-cap: "Negative Binomial Time-Series Models comparison"
# Compare Models
aicc=AIC(pois_model, nb_model, nb_seasonal, nb_offset, nb_glmm, nb_lmixed)

knitr::kable(round(aicc, 3))

```

### Model Diagnostics

```{r}
library(DHARMa)
# Check for dispersion, Zero inflation, residual patterns
# simulate residuals
sim_resi <- simulateResiduals(nb_lmixed)
plot(sim_resi)
```

```{r}
#library(DHARMa)
# Test temporal autocorrelation
#testTemporalAutocorrelation(
 # sim_resi,
#time = model_data$unique(time),
#  group = model_data$site_name
#)
```

#### Add Laged Outcome

Lag Effect (lag = 0.117, p \< 0.001)

-   For every one-unit increase in the previous month's cases, the expected current case count increases by about 12.4%, holding other variables constant.

Time Trend (time = 0.031, p \< 0.001)

-   Each additional month is associated with about a 3.2% increase in expected cases, after adjusting for seasonality, lag effects, and site-level variation.

Seasonal effects (Reference month= month1)

-   Cases are about 57% higher in month 4 than in the reference month.

Site random effect (Variance = 0.131, SD = 0.362)

-   Indicates meaningful heterogeneity among sites.

-   Some sites consistently experience higher or lower disease burdens than others, even after accounting for time, lag effects, and seasonality.

Dispersion Parameter (Dispersion = 16.3)

-   overdispersion exists, but the Negative Binomial model is explicitly accounting for it.

```{r}
#| warning: false
#| message: false
#| echo: false
model_data <- model_data %>%
  arrange(site_name, month) %>%
  group_by(site_name) %>%
  mutate(lag_cases = lag(positive_cases))
```

```{r}
#| warning: false
#| message: false
#| echo: false
# Negative Binomial Mixed Model
library(glmmTMB)
nb_lmixed1 = glmmTMB(
  positive_cases ~ + lag_cases + time + factor(month_num) + (1 | site_name),
  family = nbinom2,
  data = model_data
)

summary(nb_lmixed1)
```

```{r}
AIC(nb_lmixed, nb_lmixed1)
```

```{r}
#| warning: false
#| message: false
#| echo: false
# predict expected values
#model_data$mu <- predict(nb_lmixed1, type = "response")
# plot(model_data$mu)
# compute residual risk
# residual <- model_data$positive_cases - model_data$mu
# flag anomalies
# model_data$alert <- model_data$positive_cases >
#  qnbinom(0.95, mu = model_data$mu, size = theta)
```
